\newpage
\section{LITERATURE REVIEW}

Wireless communication has evolved significantly since the pioneering experiments of Marconi in the early 1900s, becoming a fundamental component of modern communication systems~\cite{ref3}. 
Advancements in wireless networks have been driven by the increasing demand for mobility, higher data rates, and efficient spectrum utilization~\cite{ref4}. 
Antennas remain one of the most critical components influencing communication range, throughput, and interference characteristics~\cite{ref5}. 
Modern wireless systems employ a wide variety of technologies including WiFi (IEEE 802.11), Bluetooth, NFC, LTE, 5G, and LoRa, each optimized for different ranges, data rates, and energy requirements. 
Despite these advances, reliable long-distance communication continues to face challenges due to signal attenuation, interference, and environmental dynamics.

Antennas are broadly classified as omnidirectional and directional. 
Omnidirectional antennas provide wide coverage but suffer from lower gain and higher susceptibility to interference. 
In contrast, directional antennas concentrate energy in specific directions, offering higher gain, extended range, reduced interference, and improved energy efficiency~\cite{ref6}. 
Directional antennas are therefore widely adopted in long-range networks such as Wireless Sensor Networks (WSNs), UAV communication systems, IoT deployments, and mobile ad-hoc networks. 
However, their performance depends critically on accurate alignment between transmitter and receiver. 
For directional antennas to function effectively, precise orientation is essential.

Traditional antenna alignment methods rely on GPS-based coordinate tracking or signal tracking control methods (TCM)~\cite{ref7,ref8}. 
While these approaches can achieve accurate alignment, they introduce additional hardware cost due to GPS modules and increase system complexity. 
Furthermore, GPS-based methods suffer from signal instability in indoor or obstructed environments, limiting their reliability for real-time operation in practical deployments. 
Control-theoretic approaches such as imitation human intelligence control~\cite{ref9} and servo tracking systems~\cite{ref10} provide improved automation but typically depend on accurate system modelling and parameter tuning. 
This increases implementation complexity and reduces suitability for low-cost embedded systems where computational and sensing resources are limited.

In scenarios where position information is unavailable or unreliable, antenna alignment must be performed using signal feedback, particularly the Received Signal Strength Indicator (RSSI). 
RSSI-based alignment enables orientation control without explicit knowledge of node positions, reducing hardware requirements and system cost~\cite{ref11,ref12}. 
However, early RSSI-based approaches rely on exhaustive scanning of antenna orientations to locate the direction that maximizes signal strength. 
Although simple and low-cost, exhaustive scanning is computationally and mechanically inefficient, resulting in high convergence time and poor real-time feasibility, especially in dynamic environments.

More advanced RSSI-based methods incorporate estimation and control techniques such as Kalman filtering, unscented Kalman filters, and fuzzy logic controllers~\cite{ref13}. 
These approaches improve robustness by modelling signal variations, but they introduce additional computational overhead and often require auxiliary sensors or prior system knowledge. 
As a result, hardware complexity increases and embedded suitability decreases, particularly for microcontroller-based implementations.

Autonomous antenna systems capable of self-orientation have also been explored~\cite{ref14}. 
These systems demonstrate the feasibility of low-cost mechanical alignment using RSSI feedback and servo-based actuation. 
However, their reliance on heuristic search strategies limits adaptability in noisy or dynamic environments, and convergence speed remains a concern for real-time applications.

A notable control-theoretic approach was introduced by Koru et al.~\cite{ref11}, who developed a nonlinear distributed RSSI feedback control law for GPS-free alignment of UAV-mounted directional antennas. 
While the method provides strong theoretical guarantees and eliminates the need for GPS, it requires perturbation-based estimation of RSSI gradients. 
This introduces additional measurement overhead and sensitivity to noise, increasing computational complexity and limiting real-time feasibility on resource-constrained embedded hardware.

Recent research has explored the application of machine learning (ML) to antenna alignment problems where analytical models are difficult to obtain. 
Supervised learning approaches using MLP, CNN, and LSTM architectures have demonstrated high prediction accuracy for antenna alignment based on S-parameter measurements~\cite{ref15}. 
However, these methods require large labelled datasets, offline training, and controlled data acquisition environments, which significantly increase system cost and limit their applicability for real-time embedded deployment.

Similarly, deep reinforcement learning approaches such as Deep Q-Networks (DQN) have been applied to beam alignment in mmWave and 5G systems~\cite{ref17}. 
These methods are effective in high-dimensional and dynamic environments but rely on neural network inference, which requires substantial memory and computational resources. 
Such requirements exceed the capabilities of microcontroller-class devices and make these approaches unsuitable for low-cost embedded systems without hardware acceleration.

Multi-agent reinforcement learning techniques have also been investigated for antenna tilt optimization in cellular networks~\cite{ref18}. 
While these methods enable cooperative optimization across multiple nodes, they assume network-level coordination and significant computational infrastructure, resulting in high system complexity and limited applicability to standalone embedded systems.

Overall, existing approaches present clear trade-offs between performance and implementation feasibility. 
Control-theoretic and estimation-based methods offer strong theoretical performance but require accurate models and higher computational complexity. 
Exhaustive scanning methods are simple and low-cost but lack real-time efficiency. 
Supervised and deep learning-based methods provide improved accuracy but demand large datasets and significant computational resources. 
Additionally, many solutions focus on antenna arrays and beamforming techniques, which increase hardware cost and complexity compared to mechanically steerable single-antenna systems.

\newpage
\setlength{\LTpre}{1cm}   % no space before table
\setlength{\LTpost}{1cm}  % no space after table
\renewcommand{\arraystretch}{1.4}
\setlength{\tabcolsep}{8pt}

\begin{longtable}{|
    >{\raggedright\arraybackslash}p{4.4cm} |
    >{\raggedright\arraybackslash}p{4.4cm} |
    >{\raggedright\arraybackslash}p{4.4cm} |}

\caption{Summary of Literature Review}
\label{tab:literature-review} \\
\hline
\multicolumn{1}{|c|}{\parbox[c][2em][c]{4cm}{\centering \textbf{Title}}} &
\multicolumn{1}{c|}{\parbox[c][2em][c]{4.5cm}{\centering \textbf{Methodology}}} &
\multicolumn{1}{c|}{\parbox[c][2em][c]{4.5cm}{\centering \textbf{Key Inferences}}} \\
\hline

\textit{RSSI-Based Distributed Control to Align Directional Antenna Pairs for UAV Communication} (2024) \newline
Y. Wan et al.
&
Control-theoretic approach using nonlinear static state feedback. RSSI partial derivatives estimated via antenna motion.
&
Demonstrates GPS-free RSSI-based antenna alignment with provable convergence. Strong theory, but requires accurate derivative estimation, difficult on low-cost embedded hardware.
\\ \hline

\textit{Received Signal Strength Indicator-Based Decentralised Control for Robust Long-Range Aerial Networking Using Directional Antennas} (2017) \newline
Yan et al.
&
Fusion of RSSI and GPS using Unscented Kalman Filter and fuzzy logic.
&
Improves robustness but depends on additional sensors and antenna models. Not suitable for minimal hardware systems.
\\ \hline

\textit{Self-Orientation of Directional Antennas Assisted by Mobile Robots} (2012) \newline
Min et al.
&
Pattern-based RSSI scanning using pan-tilt antenna mounts.
&
Validates RSSI-only alignment, but exhaustive scanning is slow and unsuitable for real-time systems.
\\ \hline

\textit{SmartAntenna: Enhancing Wireless Range with Autonomous Orientation} (2024) \newline
Swann et al.
&
Servo-mounted antenna using RSSI feedback and heuristic search.
&
Demonstrates low-cost feasibility, but performance degrades in noisy environments and lacks adaptive learning.
\\ \hline

\textit{Comparative Analysis of Machine Learning Algorithms for Antenna Alignments} (2024) \newline
Shakya et al.
&
Supervised ML (MLP, CNN, LSTM) trained on S-parameters.
&
High accuracy in controlled environments, but requires large datasets and offline training.
\\ \hline

\textit{Learning-Based Beam Alignment for Uplink mmWave UAVs} (2023) \newline
Susarla et al.
&
Deep Q-Network for beam alignment using antenna arrays.
&
Effective in mmWave systems, but unsuitable for single-antenna, low-cost embedded platforms.
\\ \hline

\end{longtable}


This analysis highlights a clear research gap for a low-cost, RSSI-driven antenna alignment system that can operate in real time on resource-constrained embedded hardware. 
The proposed approach addresses this gap by combining simple hardware, RSSI-only feedback, and a lightweight reinforcement learning framework. 
Unlike existing methods, it avoids dependence on GPS, complex modelling, large datasets, and high computational overhead. 
As a result, it achieves a more practical balance between cost, hardware complexity, real-time feasibility, and embedded suitability, while maintaining effective alignment performance.
