\subsection{Experimental Setup and Evaluation}

This section defines the experimental setup used to evaluate the autonomous antenna
alignment system, the metrics considered for assessment, and the protocol used to
compare the RL-based controller with the baseline methods. The goal is to ensure that
all algorithms are tested under the same initial conditions, actuator limits, and RSSI
measurement pipeline.

\subsubsection{Experimental Environment}

The system is tested in two primary environments to assess behavior under varying
propagation conditions:

\begin{itemize}
    \item \textbf{Indoor environment:} Laboratory or corridor settings with
    reflective surfaces such as walls and metallic objects. These conditions introduce
    multipath propagation and noticeable RSSI fluctuation.
    \item \textbf{Outdoor environment:} Open or semi-open areas with transmitter and
    receiver separation in the range of 10--30~m during practical trials. These
    conditions provide stronger line-of-sight behavior while still allowing mild
    obstruction scenarios.
\end{itemize}

\noindent Experiments are conducted under the following conditions:
\begin{itemize}
    \item The transmitter orientation and transmit power are kept fixed throughout a
    trial set.
    \item The receiver starts from a deliberately misaligned orientation selected from
    the same predefined set for all compared algorithms.
    \item Practical evaluation is carried out under static conditions without
    intentional disturbance so that the compared methods can be assessed under the same
    controlled setup.
    \item Multiple trials are repeated for each practical scenario.
\end{itemize}

\subsubsection{Evaluation Metrics}

To enable quantitative and fair comparison of RL-based alignment against baseline
methods, the following metrics are recorded:

\textbf{Received Signal Strength Indicator (RSSI):} 
The primary metric for assessing link quality. For each decision step, the reported
RSSI is the average of 20 consecutive packet-level measurements acquired with a
10~ms inter-sample delay. This metric is meaningful
only if it improves the received signal level observed at the wireless receiver.
Moreover, RSSI is the direct feedback signal available to all compared algorithms, so it
serves as the most fundamental common basis for comparison.
\begin{equation}
RSSI_{\text{avg}}(k) = \frac{1}{N}\sum_{i=1}^{N} RSSI_i(k), \qquad N = 20
\end{equation}

\textbf{Convergence Time ($T_{conv}$):} 
Time required for the algorithm to reach a stable alignment decision in the practical
control loop. In the deployed firmware, convergence is declared when the absolute RSSI
improvement remains below 0.8~dB for four consecutive control cycles, or when the
maximum allowed number of decision steps is reached.
The same convergence rule and runtime stopping logic are applied across the compared
methods during practical evaluation so that the reported convergence times remain
directly comparable.
\begin{equation}
\left| RSSI_{\text{avg}}(k) - RSSI_{\text{avg}}(k-1) \right| < 0.8~\text{dB},
\quad \forall k \in [k_c, k_c+3]
\end{equation}
\begin{equation}
T_{conv} = t(k_c)
\end{equation}

The 0.8~dB threshold was selected as a practical dead-band for the final hardware
system. It is small enough to indicate that the controller has entered a near-stationary
high-RSSI region, but large enough to avoid treating packet-level RSSI fluctuation or
actuator settling noise as meaningful improvement. This metric is included because the
alignment system is intended for practical use, where fast recovery from misalignment is
as important as final signal quality.

\textbf{Movement Count:} 
The total number of commanded pan and tilt movements executed before convergence.
Movement count is used as a practical proxy for actuation effort because lower motion
generally implies lower mechanical wear, lower energy use, and faster physical
alignment.
\begin{equation}
M = N_{\text{pan}} + N_{\text{tilt}}
\end{equation}

\textbf{Success Rate:} 
The percentage of repeated trials in which a method converges to a stable high-RSSI
solution within the allowed search procedure. This metric is included to distinguish fast
but unreliable methods from those that remain consistent across repeated runs.
\begin{equation}
SR(\%) = \frac{N_{\text{success}}}{N_{\text{trials}}} \times 100
\end{equation}

\textbf{RSSI Variability:} 
The spread of final RSSI values across repeated trials, reported in the results chapter
using the standard deviation of final RSSI. Lower variability indicates better
repeatability and greater resistance to environmental and measurement noise.
\begin{equation}
\sigma_{\text{RSSI}} = \sqrt{\frac{1}{N_{\text{trials}}}\sum_{j=1}^{N_{\text{trials}}}
\left(RSSI_{\text{final},j} - \overline{RSSI}_{\text{final}}\right)^2}
\end{equation}

\subsubsection{Experimental Protocol}

All experiments follow the same protocol to ensure reproducibility:

\begin{itemize}
    \item Select an initial azimuth and tilt pair from a predefined list of
    misaligned starting states.
    \item Run each algorithm from the same starting state within the same environment.
    \item Use identical runtime parameters for RSSI sampling, actuator limits, and
    settling delay across all algorithms.
    \item Record timestamps, averaged RSSI, orientation indices, chosen actions, and
    convergence status for every control cycle.
    \item Repeat each practical scenario 10 times for each compared method.
    \item Compare the RL-based method with the baseline approaches: exhaustive angular
    scan and RSSI-based hill-climbing.
\end{itemize}

\subsubsection{Statistical Analysis}

For the reported metrics, averages across repeated trials are used as the primary basis
for comparison, with standard deviation reported where it helps indicate repeatability.
Because the compared algorithms are run from matched starting conditions and shared
runtime settings, the resulting analysis supports a fair practical comparison under the
tested operating scenarios.
