\subsection{RSSI Measurement and Signal Conditioning}

The performance of the autonomous antenna alignment system relies heavily on
consistent measurement of received signal strength. Received Signal Strength
Indicator (RSSI) is a hardware-reported metric representing the power of incoming
radio packets, typically expressed in dBm. Higher RSSI values correspond to stronger
received signals and therefore provide a practical feedback signal for antenna
alignment.

Due to multipath propagation, fading, packet-to-packet variation, and environmental
interference, raw RSSI values are too noisy to be used directly for control decisions.
The measurement pipeline therefore applies lightweight conditioning before the signal
is used by the baseline algorithms or the RL policy.

\subsubsection{RSSI Sampling}

RSSI is obtained from the ESP32-S3 integrated Wi-Fi radio from packets received by the
directional antenna. Each RSSI batch is associated with a specific azimuth and tilt
orientation. In the practical implementation, 20 consecutive RSSI samples are
collected at each orientation with a 10~ms inter-sample delay, producing an effective
measurement window of approximately 200~ms per decision step. This configuration was
chosen to reduce short-term fluctuations while remaining short relative to actuator
movement and settling time.

\subsubsection{Noise Mitigation and Filtering}

RSSI measurements can fluctuate significantly even when the antenna orientation remains
fixed. To ensure robust control and learning, the following techniques are applied:
\begin{itemize}
    \item \textbf{Short-term averaging:} Multiple consecutive RSSI readings at the
    same orientation are averaged to reduce random noise
    \item \textbf{Dead-band trend classification:} The change between consecutive
    averaged RSSI values is classified using a small threshold so that minor
    fluctuations do not trigger unnecessary state changes
\end{itemize}

This conditioning ensures that the controller reacts to meaningful signal trends
rather than instantaneous fluctuations.

\subsubsection{State Feature Construction}

The RL agent requires a compact representation of the environment in the form of a
state vector. The state vector is constructed using:
\begin{itemize}
    \item \textbf{Antenna orientation:} Azimuth is discretized over the practical
    search range of $0^{\circ}$ to $180^{\circ}$ using $10^{\circ}$ steps, and tilt is
    discretized over $25^{\circ}$ to $75^{\circ}$ using $5^{\circ}$ steps.
    \item \textbf{RSSI trend:} The rate of change of averaged RSSI values is
    classified into three discrete states:
        \begin{itemize}
            \item \textbf{Increasing:} $\Delta RSSI > +0.8$ dB
            \item \textbf{Stable:} $-0.8 \leq \Delta RSSI \leq +0.8$ dB
            \item \textbf{Decreasing:} $\Delta RSSI < -0.8$ dB
        \end{itemize}
\end{itemize}

The resulting state vector is expressed as:
\begin{equation}
\mathbf{s} = \left( i_{\text{az}},\, i_{\text{tilt}},\, \Delta \mathrm{RSSI} \right)
\label{eq:state_vector}
\end{equation}
This representation captures both the current antenna orientation and short-term signal
behavior while keeping the state space small enough for direct Q-table deployment on
the ESP32-S3.

\subsubsection{Data Logging}

For debugging, performance evaluation, and offline analysis, RSSI measurements are
logged together with timestamps, azimuth and tilt indices, selected actions, and trial
identifiers. This logging supports later comparison of baseline and RL behavior under
matched runtime conditions.
