\newpage
\section{RESULTS AND ANALYSIS}

This chapter presents the final performance evaluation of the proposed autonomous
antenna alignment system. Unlike the mid-defence version of the report, which focused
primarily on simulation validation, the present section combines simulation findings
with practical hardware trials conducted in indoor and outdoor environments at
transmitter--receiver separations of 10~m and 20~m. The objective of this chapter is
to demonstrate that the reinforcement learning (RL) based controller not only performs
well in simulation, but also provides measurable improvements in real deployment when
compared with exhaustive scanning and hill-climbing.

\subsection{Final Experimental Campaign}

For final-defence validation, the implemented prototype was tested under four practical
scenarios: indoor 10~m, indoor 20~m, outdoor 10~m, and outdoor 20~m. These scenarios
were selected to represent both short-range and moderate-range communication conditions,
as well as channel environments with different levels of multipath and obstruction.

To maintain consistency, all practical experiments used the same angular search limits,
RSSI averaging configuration, and actuator resolution described in the implementation
chapter. Each algorithm was tested repeatedly under the same scenario so that mean
performance values could be compared directly.

\begin{table}[H]
\centering
\caption{Practical Experiment Configuration Used in Final Evaluation}
\label{tab:practical_experiment_config}
\renewcommand{\arraystretch}{1.3}
\begin{adjustbox}{width=\textwidth}
\begin{tabular}{|l|l|c|l|c|c|c|c|c|c|c|c|c|c|p{2.5cm}|}
\hline
\textbf{Experiment ID} & \textbf{Env.} & \textbf{Dist.} & \textbf{Obstacle} & \textbf{Pan Min} & \textbf{Pan Max} & \textbf{Tilt Min} & \textbf{Tilt Max} & \textbf{Pan Step} & \textbf{Tilt Step} & \textbf{Fine Step} & \textbf{Steps/ Deg.} & \textbf{Samples/ Point} & \textbf{Delay (ms)} & \textbf{Notes} \\
\hline
EXP\_EXH\_01 & Outdoor & 10~m & LOS & 0 & 180 & 25 & 75 & 10 & 5 & 2 & 5 & 20 & 10 & Baseline exhaustive scan \\
\hline
EXP\_EXH\_02 & Outdoor & 20~m & Partial foliage & 0 & 180 & 25 & 75 & 10 & 5 & 2 & 5 & 20 & 10 & Lower average RSSI due to distance \\
\hline
EXP\_EXH\_03 & Indoor & 10~m & Corridor LOS & 0 & 180 & 25 & 75 & 10 & 5 & 2 & 5 & 20 & 10 & Multipath reflections present \\
\hline
EXP\_EXH\_04 & Indoor & 20~m & Wall and corridor corner & 0 & 180 & 25 & 75 & 10 & 5 & 2 & 5 & 20 & 10 & Stronger fading and blockage \\
\hline
EXP\_HC\_01 & Outdoor & 10~m & LOS & 0 & 180 & 25 & 75 & 10 & 5 & 2 & 5 & 20 & 10 & Hill-climbing trials \\
\hline
EXP\_HC\_02 & Outdoor & 20~m & Partial foliage & 0 & 180 & 25 & 75 & 10 & 5 & 2 & 5 & 20 & 10 & Sensitive to initialization \\
\hline
EXP\_HC\_03 & Indoor & 10~m & Corridor LOS & 0 & 180 & 25 & 75 & 10 & 5 & 2 & 5 & 20 & 10 & Local maxima observed in some trials \\
\hline
EXP\_HC\_04 & Indoor & 20~m & Wall and corridor corner & 0 & 180 & 25 & 75 & 10 & 5 & 2 & 5 & 20 & 10 & Worst-case convergence scenario \\
\hline
EXP\_RL\_01 & Outdoor & 10~m & LOS & 0 & 180 & 25 & 75 & 10 & 5 & 2 & 5 & 20 & 10 & Deployed greedy Q-table \\
\hline
EXP\_RL\_02 & Outdoor & 20~m & Partial foliage & 0 & 180 & 25 & 75 & 10 & 5 & 2 & 5 & 20 & 10 & Practical recovery from weak start \\
\hline
EXP\_RL\_03 & Indoor & 10~m & Corridor LOS & 0 & 180 & 25 & 75 & 10 & 5 & 2 & 5 & 20 & 10 & Stable convergence across trials \\
\hline
EXP\_RL\_04 & Indoor & 20~m & Wall and corridor corner & 0 & 180 & 25 & 75 & 10 & 5 & 2 & 5 & 20 & 10 & Best performance under multipath \\
\hline
\end{tabular}
\end{adjustbox}
\end{table}

\subsection{Simulation Results Summary}

Simulation remained important in the final report because it provided a controlled means
of validating convergence behavior before hardware deployment. The simulation model was
used to compare the three alignment methods under identical path loss, directional gain,
and noise conditions. The results confirmed three key trends. First, exhaustive scanning
reliably identified the global maximum RSSI but required the highest number of samples
and mechanical movements. Second, hill-climbing was faster than exhaustive scanning but
its performance changed significantly with the starting orientation. Third, the RL agent
learned a policy that retained fast convergence while being less sensitive to poor
initialization.

\begin{table}[H]
\centering
\caption{Summary of Simulation Performance}
\label{tab:simulation_summary_final}
\renewcommand{\arraystretch}{1.4}
\begin{tabular}{|l|c|c|c|c|c|}
\hline
\textbf{Method} & \textbf{Best RSSI (dBm)} & \textbf{Convergence Time (s)} & \textbf{Actuator Steps} & \textbf{RSSI Samples} & \textbf{Robustness to Start} \\
\hline
Exhaustive scan & $-30.20$ & 67.68 & 945 & 18{,}900 & Very high \\
\hline
Hill-climbing & $-30.13$ & 6.51 & 42 & 840 & Low to moderate \\
\hline
RL (Q-learning) & $\mathbf{-29.82}$ & $\mathbf{5.51}$ & 100 & 2{,}000 & High \\
\hline
\end{tabular}
\end{table}

The simulation study therefore justified deploying the RL controller on hardware for
practical trials. However, simulation alone could not prove field effectiveness because
real measurements include electrical noise, small mechanical errors, and environmental
multipath not captured perfectly by the virtual model. For this reason, practical trials
were treated as the primary evidence in the final report.

\subsection{Practical Trial Results}

Each scenario was repeated 10 times for every alignment method. The final RSSI,
convergence time, number of movement commands, and convergence success rate were then
computed as the average across those trials. In the discussion below, higher RSSI,
shorter convergence time, fewer movements, and higher success rate indicate better
practical performance.

\begin{table}[H]
\centering
\caption{Average Practical Performance Across All Scenarios}
\label{tab:overall_practical_performance}
\renewcommand{\arraystretch}{1.4}
\begin{adjustbox}{width=\textwidth}
\begin{tabular}{|l|c|c|c|c|c|}
\hline
\textbf{Method} & \textbf{Avg. Final RSSI (dBm)} & \textbf{Avg. Conv. Time (s)} & \textbf{Avg. Moves} & \textbf{Success Rate (\%)} & \textbf{Post-Conv. RSSI Std. Dev. (dB)} \\
\hline
Exhaustive scan & $-48.95$ & 54.20 & 812 & 100 & 1.60 \\
\hline
Hill-climbing & $-52.73$ & 11.84 & 63 & 82.5 & 2.34 \\
\hline
RL (Q-learning) & $\mathbf{-46.55}$ & $\mathbf{8.62}$ & 57 & $\mathbf{97.5}$ & $\mathbf{1.18}$ \\
\hline
\end{tabular}
\end{adjustbox}
\end{table}

Table~\ref{tab:overall_practical_performance} shows that RL achieved the best overall
combination of link quality, speed, repeatability, and stability. Exhaustive scanning
retained a perfect success rate because it examines the whole search space, but this was
achieved at a very large time and movement cost. Hill-climbing reduced convergence time
substantially, yet it underperformed in final RSSI and success rate because some trials
terminated at locally favorable but globally inferior antenna orientations. The RL agent
provided the highest average RSSI and the lowest convergence time while keeping movement
count lower than exhaustive scanning and comparable to hill-climbing.

\subsubsection{Scenario-wise Comparison}

To understand the behavior in more detail, the performance of each method was also
analyzed separately for each environment and distance.

\begin{table}[H]
\centering
\caption{Scenario-wise Practical Results for Exhaustive Scan}
\label{tab:exhaustive_scenario_results}
\renewcommand{\arraystretch}{1.35}
\begin{adjustbox}{width=\textwidth}
\begin{tabular}{|l|c|c|c|c|c|}
\hline
\textbf{Scenario} & \textbf{Final RSSI (dBm)} & \textbf{Conv. Time (s)} & \textbf{Moves} & \textbf{Success (\%)} & \textbf{Remarks} \\
\hline
Outdoor, 10~m & $-41.8 \pm 0.9$ & $49.2 \pm 1.5$ & $812 \pm 0$ & 100 & Accurate but time-consuming \\
\hline
Outdoor, 20~m & $-47.1 \pm 1.1$ & $53.8 \pm 1.8$ & $812 \pm 0$ & 100 & Lower RSSI due to distance \\
\hline
Indoor, 10~m & $-50.4 \pm 1.3$ & $55.1 \pm 2.0$ & $812 \pm 0$ & 100 & Multipath increases variance \\
\hline
Indoor, 20~m & $-56.5 \pm 1.6$ & $58.7 \pm 2.3$ & $812 \pm 0$ & 100 & Slowest practical case \\
\hline
\end{tabular}
\end{adjustbox}
\end{table}

\begin{table}[H]
\centering
\caption{Scenario-wise Practical Results for Hill-Climbing}
\label{tab:hill_scenario_results}
\renewcommand{\arraystretch}{1.35}
\begin{adjustbox}{width=\textwidth}
\begin{tabular}{|l|c|c|c|c|c|}
\hline
\textbf{Scenario} & \textbf{Final RSSI (dBm)} & \textbf{Conv. Time (s)} & \textbf{Moves} & \textbf{Success (\%)} & \textbf{Remarks} \\
\hline
Outdoor, 10~m & $-44.9 \pm 1.8$ & $9.1 \pm 1.9$ & $51 \pm 9$ & 90 & Good when initialized near optimum \\
\hline
Outdoor, 20~m & $-50.7 \pm 2.4$ & $10.8 \pm 2.2$ & $58 \pm 11$ & 80 & More local-trap cases observed \\
\hline
Indoor, 10~m & $-54.1 \pm 2.9$ & $12.6 \pm 2.7$ & $66 \pm 12$ & 80 & Reflections mislead local search \\
\hline
Indoor, 20~m & $-61.2 \pm 3.5$ & $14.9 \pm 3.0$ & $77 \pm 13$ & 70 & Weakest performance in obstructed case \\
\hline
\end{tabular}
\end{adjustbox}
\end{table}

\begin{table}[H]
\centering
\caption{Scenario-wise Practical Results for RL-Based Alignment}
\label{tab:rl_scenario_results}
\renewcommand{\arraystretch}{1.35}
\begin{adjustbox}{width=\textwidth}
\begin{tabular}{|l|c|c|c|c|c|}
\hline
\textbf{Scenario} & \textbf{Final RSSI (dBm)} & \textbf{Conv. Time (s)} & \textbf{Moves} & \textbf{Success (\%)} & \textbf{Remarks} \\
\hline
Outdoor, 10~m & $\mathbf{-40.2 \pm 0.8}$ & $\mathbf{6.8 \pm 1.1}$ & $\mathbf{43 \pm 6}$ & 100 & Fastest and strongest link quality \\
\hline
Outdoor, 20~m & $\mathbf{-45.5 \pm 1.0}$ & $\mathbf{8.1 \pm 1.2}$ & $\mathbf{52 \pm 7}$ & 100 & Stable despite weaker received power \\
\hline
Indoor, 10~m & $\mathbf{-48.3 \pm 1.2}$ & $\mathbf{8.9 \pm 1.5}$ & $\mathbf{61 \pm 8}$ & 100 & Robust to corridor reflections \\
\hline
Indoor, 20~m & $\mathbf{-52.2 \pm 1.7}$ & $\mathbf{10.7 \pm 1.6}$ & $\mathbf{72 \pm 9}$ & 90 & Best among all methods in hardest scenario \\
\hline
\end{tabular}
\end{adjustbox}
\end{table}

The practical measurements show a consistent trend across all four scenarios. In outdoor
line-of-sight testing at 10~m, RL converged 86.2\% faster than exhaustive scanning and
25.3\% faster than hill-climbing, while also improving the final RSSI by 1.6~dB over
exhaustive scanning and 4.7~dB over hill-climbing. At 20~m outdoors, RL preserved this
advantage despite the expected reduction in received power due to distance.

The indoor scenarios were more challenging because reflected components introduced local
peaks in the RSSI surface. In these cases, hill-climbing frequently converged to local
maxima, particularly at 20~m where obstruction and multipath were both present. RL,
although not perfect in every single trial, maintained the highest success rate and the
best final RSSI among all tested methods. This behavior indicates that the learned policy
captures useful directional tendencies that generalize better than purely greedy search.

\subsubsection{Relative Improvement of RL Over Baselines}

To quantify the benefit of RL more directly, the average improvement of RL over the two
baseline methods was computed for each scenario.

\begin{table}[H]
\centering
\caption{Improvement of RL Relative to Baseline Methods}
\label{tab:rl_relative_improvement}
\renewcommand{\arraystretch}{1.35}
\begin{adjustbox}{width=\textwidth}
\begin{tabular}{|l|c|c|c|c|}
\hline
\textbf{Scenario} & \textbf{RSSI Gain over Exhaustive (dB)} & \textbf{RSSI Gain over Hill (dB)} & \textbf{Time Reduction vs Exhaustive (\%)} & \textbf{Time Reduction vs Hill (\%)} \\
\hline
Outdoor, 10~m & 1.6 & 4.7 & 86.2 & 25.3 \\
\hline
Outdoor, 20~m & 1.6 & 5.2 & 84.9 & 25.0 \\
\hline
Indoor, 10~m & 2.1 & 5.8 & 83.8 & 29.4 \\
\hline
Indoor, 20~m & 4.3 & 9.0 & 81.8 & 28.2 \\
\hline
\end{tabular}
\end{adjustbox}
\end{table}

The largest practical gain of RL occurred in the most difficult indoor 20~m case, where
it improved final RSSI by 4.3~dB compared with exhaustive scanning and by 9.0~dB compared
with hill-climbing. This result is significant because it suggests that the RL controller
is not merely faster; it is also better at selecting stable high-gain orientations in
complex propagation environments.

\subsection{Analysis of Practical Behavior}

The observed results can be explained by the search behavior of each algorithm. The
exhaustive method is reliable because it samples the full angle space, but this same
property makes it inefficient and unsuitable for repeated realignment. Hill-climbing is
computationally simple and fast, but it assumes that local RSSI improvements always point
toward the global optimum. In indoor measurements this assumption is often violated due
to multipath, leading to premature convergence.

The RL agent performs better because its decision policy is shaped by repeated exposure to
noisy and suboptimal states during training. Instead of reacting to a single local RSSI
increase, it uses a learned value estimate to choose actions that are more likely to lead
to sustained future reward. This gives the method greater resilience against temporary
signal fluctuations and misleading local peaks. The low post-convergence RSSI variation
reported in Table~\ref{tab:overall_practical_performance} further indicates that the RL
controller settles at more stable orientations than hill-climbing.

Another important outcome is that the practical trials broadly agree with the simulation
study. In both cases, exhaustive scanning provided a high-confidence reference, hill-
climbing offered faster but less reliable convergence, and RL achieved the best balance
between speed, accuracy, and robustness. This consistency strengthens the validity of the
proposed approach and supports the use of simulation-assisted RL design for embedded
wireless alignment systems.

\subsection{Discussion}

From an engineering perspective, the final evaluation demonstrates that the proposed RL-
based alignment system is feasible for embedded deployment and beneficial in practical
operation. The policy remains compact enough for direct storage on the ESP32-S3, the
runtime decision process requires only lightweight table lookup, and the measured gains in
RSSI and convergence speed justify the added effort of offline training.

The practical results also show that the RL controller is especially useful in scenarios
where the environment is not perfectly smooth, such as indoor corridors and partially
obstructed outdoor links. These are precisely the scenarios where a conventional greedy
search method becomes unreliable. Consequently, the final report concludes that the RL-
based controller is the most suitable of the evaluated methods for autonomous directional
antenna alignment under the constraints of low-cost embedded hardware.
