\subsection{Final Experimental Setup}

The final evaluation was carried out in four practical scenarios selected to represent
both outdoor and indoor operation under the practical conditions used in the final
prototype. Outdoor testing was carried out at 20~m and 30~m, while indoor testing was
performed at 10~m under both line-of-sight and obstructed conditions. All three
alignment methods were tested using the same search limits, RSSI acquisition settings,
and mechanical resolution so that the measured differences reflect the control
strategy rather than a change in experimental setup.

Each practical scenario was repeated 10 times for every alignment method. The final
RSSI, convergence time, movement count, and success rate reported in the following
subsections are averages over those repeated trials.

\setlength{\tabcolsep}{4pt}

\vspace{0.5cm}
\begin{table}[H]
\centering
\caption{Shared Runtime Parameters for Practical Trials}
\label{tab:practical_runtime_results}
\renewcommand{\arraystretch}{1.25}
\begin{adjustbox}{max width=\textwidth}
\begin{tabular}{|
>{\raggedright\arraybackslash}p{1.7cm}|
>{\raggedright\arraybackslash}p{1.7cm}|
>{\raggedright\arraybackslash}p{1.4cm}|
>{\raggedright\arraybackslash}p{1.4cm}|
>{\raggedright\arraybackslash}p{1.4cm}|
>{\raggedright\arraybackslash}p{1.8cm}|
>{\raggedright\arraybackslash}p{2.0cm}|
>{\raggedright\arraybackslash}p{1.8cm}|}
\hline
\textbf{Pan Range} & \textbf{Tilt Range} & \textbf{Pan Step} & \textbf{Tilt Step} & \textbf{Fine Step} & \textbf{Steps/ Deg.} & \textbf{Samples/ Point} & \textbf{Sample Delay} \\
\hline
$0^\circ$--$180^\circ$ & $25^\circ$--$75^\circ$ & $10^\circ$ & $5^\circ$ & $2^\circ$ & 5 & 20 & 10~ms \\
\hline
\end{tabular}
\end{adjustbox}
\end{table}

\vspace{0.5cm}
\begin{table}[H]
\centering
\caption{Practical Test Scenarios Used in Final Evaluation}
\label{tab:practical_scenarios}
\renewcommand{\arraystretch}{1.25}
\begin{adjustbox}{width=\textwidth}
\begin{tabular}{|
>{\raggedright\arraybackslash}p{1.15cm}|
>{\raggedright\arraybackslash}p{1.85cm}|
>{\raggedright\arraybackslash}p{1.35cm}|
>{\raggedright\arraybackslash}p{3.85cm}|
>{\raggedright\arraybackslash}p{4.25cm}|}
\hline
\textbf{ID} & \textbf{Env.} & \textbf{Dist.} & \textbf{Channel Condition} & \textbf{Purpose} \\
\hline
S1 & Outdoor & 20~m & Clear LOS & Reference case with minimal obstruction and strong received power \\
\hline
S2 & Outdoor & 30~m & Partial foliage & Tests longer range and mild attenuation in a mostly open channel \\
\hline
S3 & Indoor & 10~m & Corridor LOS with reflections & Evaluates sensitivity to multipath in a confined environment \\
\hline
S4 & Indoor & 10~m & Wall and corridor corner & Hardest indoor case with obstruction, fading, and multipath peaks \\
\hline
\end{tabular}
\end{adjustbox}
\end{table}

\subsection{Practical Trial Results}

Practical trials provide the primary evidence for the final project because they
include real RF noise, actuator delay, and mechanical imperfections that are not fully
captured by the simulator.

\vspace{0.5cm}
\begin{table}[H]
\centering
\caption{Average Practical Performance Across All Scenarios}
\label{tab:overall_practical_performance}
\renewcommand{\arraystretch}{1.25}
\begin{adjustbox}{width=\textwidth}
\begin{tabular}{|
>{\raggedright\arraybackslash}p{2.35cm}|
>{\raggedright\arraybackslash}p{2.0cm}|
>{\raggedright\arraybackslash}p{1.9cm}|
>{\raggedright\arraybackslash}p{1.6cm}|
>{\raggedright\arraybackslash}p{1.75cm}|
>{\raggedright\arraybackslash}p{2.45cm}|}
\hline
\textbf{Method} & \textbf{Avg. RSSI (dBm)} & \textbf{Avg. Time (s)} & \textbf{Avg. Moves} & \textbf{Success (\%)} & \textbf{RSSI Std. Dev. (dB)} \\
\hline
Exhaustive scan & $-48.95$ & 54.20 & 812 & 100 & 1.60 \\
\hline
Hill-climbing & $-52.73$ & 11.84 & 63 & 80 & 2.34 \\
\hline
RL (Q-learning) & $-49.55$ & $\mathbf{9.62}$ & $\mathbf{57}$ & $\mathbf{95}$ & $\mathbf{1.28}$ \\
\hline
\end{tabular}
\end{adjustbox}
\end{table}

The overall averages show that RL achieved much faster convergence than exhaustive
scan while maintaining a final RSSI close to the exhaustive reference and clearly
better than hill-climbing. Exhaustive scan retained the best final RSSI and perfect
success rate because it explores the full search space, but that reliability came at
very high time and movement cost. Hill-climbing was faster than exhaustive scan,
yet its lower success rate and weaker final RSSI show that it was more easily trapped
by local RSSI peaks.

\subsubsection{Scenario-wise Practical Comparison}

\vspace{0.5cm}
\renewcommand{\arraystretch}{1.25}
\setlength{\LTpre}{8pt}
\setlength{\LTpost}{8pt}
\begin{longtable}{|
>{\raggedright\arraybackslash}p{4.2cm}|
>{\raggedright\arraybackslash}p{3.5cm}|
>{\raggedright\arraybackslash}p{3cm}|
>{\raggedright\arraybackslash}p{3cm}|}
\caption{Scenario-wise Final RSSI Comparison}
\label{tab:scenario_rssi_comparison}\\
\hline
\textbf{Scenario} & \textbf{Exhaustive (dBm)} & \textbf{Hill-Climbing (dBm)} & \textbf{RL (dBm)} \\
\hline
\endfirsthead
\hline
\textbf{Scenario} & \textbf{Exhaustive (dBm)} & \textbf{Hill-Climbing (dBm)} & \textbf{RL (dBm)} \\
\hline
\endhead
Outdoor, 20~m & $-41.8 \pm 0.9$ & $-44.9 \pm 1.8$ & $-42.1 \pm 0.9$ \\
\hline
Outdoor, 30~m & $-47.1 \pm 1.1$ & $-50.7 \pm 2.4$ & $-47.6 \pm 1.1$ \\
\hline
Indoor, 10~m (LOS) & $-50.4 \pm 1.3$ & $-54.1 \pm 2.9$ & $-50.9 \pm 1.3$ \\
\hline
Indoor, 10~m (obstructed) & $-56.5 \pm 1.6$ & $-61.2 \pm 3.5$ & $-57.6 \pm 1.8$ \\
\hline
\end{longtable}

\vspace{0.5cm}
\renewcommand{\arraystretch}{1.25}
\begin{longtable}{|
>{\raggedright\arraybackslash}p{4.2cm}|
>{\raggedright\arraybackslash}p{3.0cm}|
>{\raggedright\arraybackslash}p{3.0cm}|
>{\raggedright\arraybackslash}p{3.0cm}|}
\caption{Scenario-wise Convergence Time and Success Rate}
\label{tab:scenario_time_success}\\
\hline
\textbf{Scenario} & \textbf{Exhaustive} & \textbf{Hill-Climbing} & \textbf{RL} \\
\hline
\endfirsthead
\hline
\textbf{Scenario} & \textbf{Exhaustive} & \textbf{Hill-Climbing} & \textbf{RL} \\
\hline
\endhead
Outdoor, 20~m & $49.2 \pm 1.5$ s, 100\% & $9.1 \pm 1.9$ s, 90\% & $\mathbf{7.8 \pm 1.2}$ s, 100\% \\
\hline
Outdoor, 30~m & $53.8 \pm 1.8$ s, 100\% & $10.8 \pm 2.2$ s, 80\% & $\mathbf{9.0 \pm 1.3}$ s, 100\% \\
\hline
Indoor, 10~m (LOS) & $55.1 \pm 2.0$ s, 100\% & $12.6 \pm 2.7$ s, 80\% & $\mathbf{10.1 \pm 1.6}$ s, 90\% \\
\hline
Indoor, 10~m (obstructed) & $58.7 \pm 2.3$ s, 100\% & $14.9 \pm 3.0$ s, 70\% & $\mathbf{11.6 \pm 1.8}$ s, 90\% \\
\hline
\end{longtable}

\vspace{0.5cm}
\begin{figure}[H]
\centering
\begin{tikzpicture}
\begin{axis}[
    ybar,
    bar width=7pt,
    width=\textwidth,
    height=7.2cm,
    ymin=-65,
    ymax=-35,
    ylabel={Final RSSI (dBm)},
    symbolic x coords={Out-20,Out-30,In-10L,In-10O},
    xtick=data,
    enlarge x limits=0.18,
    legend style={at={(0.5,1.05)}, anchor=south, legend columns=3},
    grid=major
]
\addplot[fill=gray!45] coordinates {(Out-20,-41.8) (Out-30,-47.1) (In-10L,-50.4) (In-10O,-56.5)};
\addplot[fill=orange!60] coordinates {(Out-20,-44.9) (Out-30,-50.7) (In-10L,-54.1) (In-10O,-61.2)};
\addplot[fill=blue!50] coordinates {(Out-20,-42.1) (Out-30,-47.6) (In-10L,-50.9) (In-10O,-57.6)};
\legend{Exhaustive,Hill-Climbing,RL}
\end{axis}
\end{tikzpicture}
\vspace{0.15cm}
\caption{Scenario-wise final RSSI comparison in practical trials}
\label{fig:practical_rssi_chart}
\end{figure}

\begin{figure}[H]
\centering
\begin{tikzpicture}
\begin{axis}[
    ybar,
    bar width=7pt,
    width=\textwidth,
    height=7.2cm,
    ymin=0,
    ylabel={Convergence Time (s)},
    symbolic x coords={Out-20,Out-30,In-10L,In-10O},
    xtick=data,
    enlarge x limits=0.18,
    legend style={at={(0.5,1.05)}, anchor=south, legend columns=3},
    grid=major
]
\addplot[fill=gray!45] coordinates {(Out-20,49.2) (Out-30,53.8) (In-10L,55.1) (In-10O,58.7)};
\addplot[fill=orange!60] coordinates {(Out-20,9.1) (Out-30,10.8) (In-10L,12.6) (In-10O,14.9)};
\addplot[fill=blue!50] coordinates {(Out-20,7.8) (Out-30,9.0) (In-10L,10.1) (In-10O,11.6)};
\legend{Exhaustive,Hill-Climbing,RL}
\end{axis}
\end{tikzpicture}
\vspace{0.15cm}
\caption{Scenario-wise convergence time comparison in practical trials}
\label{fig:practical_time_chart}
\end{figure}

Figures~\ref{fig:practical_rssi_chart} and~\ref{fig:practical_time_chart} show a
consistent practical trend across all four scenarios. Exhaustive scan achieved the
best final RSSI in each case, as expected from a full search of the sampled
orientation space. RL converged much faster than exhaustive scan while reaching a
final RSSI close to the exhaustive reference and still outperforming hill-climbing in
every scenario. The performance gap widened in the obstructed indoor 10~m case,
where the wireless channel was least smooth and hill-climbing was most vulnerable to
local maxima.

\vspace{0.5cm}
\begin{table}[H]
\centering
\fontsize{12}{14}\selectfont
\caption{Improvement of RL Relative to Baseline Methods}
\label{tab:rl_relative_improvement}
\renewcommand{\arraystretch}{1.25}
\begin{adjustbox}{max width=\textwidth}
\begin{tabular}{|
>{\raggedright\arraybackslash}p{2.55cm}|
>{\raggedright\arraybackslash}p{2.2cm}|
>{\raggedright\arraybackslash}p{2.2cm}|
>{\raggedright\arraybackslash}p{2.4cm}|
>{\raggedright\arraybackslash}p{2.4cm}|}
\hline
\textbf{Scenario} & \textbf{RSSI Gain vs Exh.} & \textbf{RSSI Gain vs Hill} & \textbf{Time Reduction vs Exh.} & \textbf{Time Reduction vs Hill} \\
\hline
Outdoor, 20~m & $-0.3$~dB & 2.8~dB & 84.1\% & 14.3\% \\
\hline
Outdoor, 30~m & $-0.5$~dB & 3.1~dB & 83.3\% & 16.7\% \\
\hline
Indoor, 10~m (LOS) & $-0.5$~dB & 3.2~dB & 81.7\% & 19.8\% \\
\hline
Indoor, 10~m (obstructed) & $-1.1$~dB & 3.6~dB & 80.2\% & 22.1\% \\
\hline
\end{tabular}
\end{adjustbox}
\end{table}

The improvement table confirms that RL does not surpass exhaustive scan in final
RSSI, which is the realistic outcome for a complete search. Its practical advantage
is instead the much lower convergence time, with only a small RSSI sacrifice relative
to exhaustive scan and a clear gain over hill-climbing. This tradeoff is especially
valuable in the harder indoor scenarios, where fast but purely greedy search becomes
less reliable.

\subsection{Validation Through Theory, Simulation, and Practical Output}

The measured outputs agree well with the expected qualitative behavior of the three
algorithms. Table~\ref{tab:theory_sim_practical_validation} summarizes this validation.

\vspace{0.5cm}
\renewcommand{\arraystretch}{1.25}
\begin{longtable}{|
>{\raggedright\arraybackslash}p{2.35cm}|
>{\raggedright\arraybackslash}p{1.35cm}|
>{\raggedright\arraybackslash}p{2.75cm}|
>{\raggedright\arraybackslash}p{1.65cm}|
>{\raggedright\arraybackslash}p{1.35cm}|
>{\raggedright\arraybackslash}p{1.75cm}|
>{\raggedright\arraybackslash}p{1.65cm}|}
\caption{Simulation-to-Practical Deviation by Method}
\label{tab:simulation_practical_deviation}\\
\hline
\textbf{Method} & \textbf{Sim. RSSI} & \textbf{Prac. Avg. RSSI} & \textbf{$\Delta$ RSSI} & \textbf{Sim. Time} & \textbf{Prac. Avg. Time} & \textbf{$\Delta$ Time} \\
\hline
\endfirsthead
\hline
\textbf{Method} & \textbf{Sim. RSSI} & \textbf{Prac. Avg. RSSI} & \textbf{$\Delta$ RSSI} & \textbf{Sim. Time} & \textbf{Prac. Avg. Time} & \textbf{$\Delta$ Time} \\
\hline
\endhead
Exhaustive scan & $-30.20$ & $-48.95$ & $-18.75$ dB & 67.68 s & 54.20 s & $-19.9\%$ \\
\hline
Hill-climbing & $-30.13$ & $-52.73$ & $-22.60$ dB & 6.51 s & 11.84 s & $+81.9\%$ \\
\hline
RL (Q-learning) & $-29.82$ & $-49.55$ & $-19.73$ dB & 5.51 s & 9.62 s & $+74.6\%$ \\
\hline
\end{longtable}


The absolute RSSI values in practice are lower than in simulation for all methods,
which is expected because the simulator does not capture every source of path loss,
connector loss, mechanical misalignment, and environmental attenuation. RL still
transfers better than hill-climbing in the sense that it preserves a stronger final
RSSI and better repeatability, but the practical measurements also show that its
performance remains below the exhaustive reference, which is consistent with a
time-efficient search rather than a full search.


\vspace{0.5cm}
\renewcommand{\arraystretch}{1.25}
\begin{longtable}{|
>{\raggedright\arraybackslash}p{3.85cm}|
>{\raggedright\arraybackslash}p{4.25cm}|
>{\raggedright\arraybackslash}p{2.95cm}|
>{\raggedright\arraybackslash}p{3.0cm}|}
\caption{Validation of Observed Behavior Against Expected Theory}
\label{tab:theory_sim_practical_validation}\\
\hline
\textbf{Expected Behavior} & \textbf{Theory} & \textbf{Simulation Output} & \textbf{Practical Output} \\
\hline
\endfirsthead
\hline
\textbf{Expected Behavior} & \textbf{Theory} & \textbf{Simulation Output} & \textbf{Practical Output} \\
\hline
\endhead
Exhaustive scan should be reliable but slow & Full search should locate the best sampled orientation at the cost of many measurements & Highest convergence time and largest step count & 100\% success in all scenarios, but by far the slowest method \\
\hline
Hill-climbing should be fast but locally sensitive & Greedy local search can terminate at suboptimal peaks & Fast convergence but low-to-moderate robustness & Short practical convergence time but reduced success rate and weaker final RSSI indoors \\
\hline
RL should balance speed and robustness & Learned value estimates should guide the search beyond immediate local RSSI gradients & Fastest simulated convergence and best final RSSI & Near-exhaustive average RSSI, faster convergence than exhaustive scan, and strong repeatability \\
\hline
Indoor channels should be harder than outdoor channels & Multipath and obstruction distort the RSSI surface & Not fully captured in idealized form & Lower final RSSI and longer convergence time for all methods, especially hill-climbing \\
\hline
\end{longtable}

The comparison shows that the practical measurements did not contradict the theoretical
or simulated expectations. Instead, they preserved the same ranking of algorithmic
behavior while introducing additional degradation associated with real hardware and RF
conditions.

\subsection{Error Analysis}

The difference between expected and measured outputs can be traced to several practical
non-idealities in the hardware platform and test environment.

\vspace{0.5cm}
\renewcommand{\arraystretch}{1.25}
\begin{longtable}{|
>{\raggedright\arraybackslash}p{3cm}|
>{\raggedright\arraybackslash}p{3.2cm}|
>{\raggedright\arraybackslash}p{3cm}|
>{\raggedright\arraybackslash}p{4.45cm}|}
\caption{Sources of Error and Their Impact on Results}
\label{tab:error_analysis}\\
\hline
\textbf{Source of Error} & \textbf{Effect on Output} & \textbf{Observed Evidence} & \textbf{Mitigation Used} \\
\hline
\endfirsthead
\hline
\textbf{Source of Error} & \textbf{Effect on Output} & \textbf{Observed Evidence} & \textbf{Mitigation Used} \\
\hline
\endhead
Multipath propagation and obstruction & Creates local RSSI peaks and fluctuating measurements & Larger variance and poorer hill-climbing performance in indoor cases & RSSI averaging, repeated trials, comparison across indoor and outdoor scenarios \\
\hline
Mechanical backlash and finite step resolution & Causes the realized angle to differ slightly from the commanded angle & Practical RSSI lower than simulated ideal values and slower settling after movement & Limited search range, calibrated step factor, mechanical settling delay \\
\hline
Slip-ring contact variation and electrical noise & Perturbs supply quality and signal stability during rotation & Trial-to-trial RSSI fluctuation and small convergence variability & Local decoupling, signal separation, short averaging window before decision making \\
\hline
Servo positioning error and step quantization & Prevents exact arrival at the continuous optimum orientation & Residual difference between exhaustive reference and final aligned angle & Discrete state/action design matched to hardware resolution \\
\hline
Simulation model simplification & Underestimates real environmental complexity & All practical RSSI values are lower than simulated values & Practical validation used as final evidence, not simulation alone \\
\hline
\end{longtable}

Among these sources, multipath and actuator non-idealities had the greatest influence
on deviation from expected values. They are also the main reason why hill-climbing
performed poorly in the indoor scenarios. The RL controller was less affected because
it does not rely only on the instantaneous local gradient of RSSI and instead follows a
policy learned from repeated interaction with noisy states.

\subsection{Overall Analysis}

From an engineering perspective, the outputs of this chapter validate the proposed
project in both simulation and practice. The compact Q-table could be deployed on the
ESP32-S3, the real system converged in all tested scenarios with high reliability, and
the RL controller consistently outperformed the two baseline methods in the combined
criteria of RSSI quality, convergence time, and robustness.

The practical results are especially important because they show that the project does
not only work under simulated conditions. Even after accounting for hardware error,
environmental attenuation, and multipath, the RL-based controller remained the most
suitable of the tested methods for autonomous directional antenna alignment on low-cost
embedded hardware.
