\newpage
\section{CONCLUSION}

This project developed an autonomous directional antenna alignment prototype based on
reinforcement learning and low-cost embedded hardware. The work combined a custom
mechanical platform, ESP32-S3-based sensing and actuation, offline Q-learning, and both
simulation and practical evaluation into a complete end-to-end system.
The final results indicate that, in the tested scenarios, the RL-based controller offered
a useful trade-off between alignment quality and convergence time. Compared with
exhaustive scanning, the RL controller converged substantially faster while achieving
final RSSI values that remained close to the exhaustive reference. Compared with
hill-climbing, the RL method showed more consistent performance across the practical
trials, especially in indoor multipath conditions and partially obstructed outdoor
conditions.

The system was developed under constraints of cost and component availability, which
influenced certain hardware design decisions, including the use of a custom PCB-based slip
ring and readily available actuators; as a result, the current prototype remains limited
in long-term reliability, signal integrity, mechanical precision, and the range of
practical conditions tested.
In addition, the RL controller depends on offline training over a discretized state space,
so its performance may reduce when operating conditions differ significantly from those
represented during training.

Overall, the project suggests that reinforcement learning is a good approach for
autonomous antenna alignment in low-cost wireless systems. The prototype and results
presented in this report provide a foundation for future work on adaptive beam steering,
intelligent wireless links, and embedded RL-driven control systems.
