\newpage
\section*{ABSTRACT}  
\phantomsection
\addcontentsline{toc}{section}{ABSTRACT}

Precise alignment of directional antennas is critical for maintaining reliable
wireless communication links, particularly in low-cost and resource-constrained
systems where phased arrays and complex beamforming hardware are impractical.
This project presents the design and implementation of an autonomous antenna
alignment system that uses a reinforcement learning-based policy to select antenna
orientations based on Received Signal Strength Indicator (RSSI) measurements. The
system comprises a fixed transmitter node and a receiver node equipped with an
ESP32-S3 microcontroller, a high-gain 2.4~GHz directional antenna, a stepper-driven
azimuth stage, and a servo-driven tilt stage. RSSI measurements are averaged and
encoded into a compact discrete state representation, which is used to query a
pre-trained tabular Q-learning policy stored on the ESP32-S3. The selected actions
command the azimuth and tilt actuators to adjust the antenna orientation in real
time. To assess the suitability of this approach for low-cost embedded deployment,
the system is designed to be evaluated against conventional baseline methods,
including fixed orientation, exhaustive angular scanning, and RSSI-based
hill-climbing. The evaluation framework considers convergence time, alignment
accuracy, link stability, and mechanical efficiency under indoor and outdoor test
conditions. Across the tested practical scenarios, the reinforcement learning controller 
achieved an average final RSSI of $-49.55$~dBm, compared to $-48.95$~dBm obtained using 
exhaustive angular scanning, indicating a marginal reduction in peak signal strength. 
However, the proposed method reduced the average convergence time from 54.20~s to 9.62~s, 
corresponding to an improvement of approximately 82.25\%.
These results indicate that the proposed method provides a useful time--accuracy trade-off
for autonomous directional antenna alignment on resource-constrained wireless
systems, rather than absolute RSSI superiority over exhaustive search.

\vspace{0.5cm}

\textit{Keywords: Directional Antenna Alignment, Embedded Systems, Policy-based Control, Reinforcement Learning, RSSI-based Optimization, Wireless Communication}
