\subsection{Reinforcement Learning Implementation}

This subsection describes how the Q-learning controller was trained, converted into an
embedded-friendly representation, and integrated into the practical alignment workflow.
The design objective was to keep the learned policy lightweight enough for execution on
the ESP32-S3 while retaining sufficient state information to handle noisy RSSI
observations and changing starting orientations.

\subsubsection{State, Action, and Reward Representation}

The deployed RL state is defined as:
\begin{equation}
s_k = \left( i_{\text{pan}},\; i_{\text{tilt}},\; \text{trend}_k \right)
\end{equation}
where the first two terms represent discretized pan and tilt indices and the third term
captures the short-term RSSI trend. This representation allows the agent to exploit both
spatial orientation and immediate signal feedback without requiring a large state space.

The action set used during training and deployment is:
\begin{equation}
\mathcal{A} = \{(+\Delta_p,0), (-\Delta_p,0), (0,+\Delta_t), (0,-\Delta_t), (0,0)\}
\end{equation}
where $\Delta_p$ and $\Delta_t$ denote the practical pan and tilt increments. The hold
action was retained so that the controller could stop moving once the alignment entered
a stable high-RSSI region.

The reward function favors signal improvement while penalizing unnecessary movement:
\begin{equation}
r_k = \Delta \text{RSSI}_k - \lambda \cdot \mathbb{I}_{\text{move}}
\end{equation}
This reward shaping improved convergence smoothness and reduced oscillatory behavior in
both simulation and practical inference.

\subsubsection{Training Configuration}

Offline training was carried out in the Python simulation environment using repeated
episodes with randomized starting states. Hyperparameters were tuned so that the agent
would converge reliably without overfitting to a single start angle.

\begin{table}[H]
\centering
\caption{Q-Learning Training Configuration}
\label{tab:q_learning_parameters}
\renewcommand{\arraystretch}{1.4}
\begin{tabular}{|p{5.2cm}|p{2.6cm}|p{5.2cm}|}
\hline
\textbf{Parameter} & \textbf{Value} & \textbf{Purpose} \\
\hline
Number of training episodes & 2500 & Provides repeated exposure to diverse initial orientations \\
\hline
Maximum steps per episode & 60 & Prevents long unproductive trajectories \\
\hline
Learning rate $\alpha$ & 0.18 & Balances adaptation speed and numerical stability \\
\hline
Discount factor $\gamma$ & 0.92 & Encourages actions with long-term RSSI improvement \\
\hline
Initial exploration rate $\epsilon$ & 1.0 & Supports broad exploration at the start of training \\
\hline
Final exploration rate & 0.05 & Preserves limited exploration near convergence \\
\hline
Exploration decay & 0.995 / episode & Gradually shifts learning toward exploitation \\
\hline
Movement penalty $\lambda$ & 0.35 & Discourages unnecessary actuator motion \\
\hline
\end{tabular}
\end{table}

\subsubsection{Q-Table Storage and Embedded Deployment}

After training convergence, the Q-table was exported from Python and converted into a
static array representation suitable for firmware inclusion. This avoided the need for
runtime file parsing on the microcontroller and ensured deterministic startup behavior.

\begin{table}[H]
\centering
\caption{Embedded RL Deployment Summary}
\label{tab:rl_deployment_summary}
\renewcommand{\arraystretch}{1.4}
\begin{tabular}{|p{4.8cm}|p{2.6cm}|p{5.3cm}|}
\hline
\textbf{Item} & \textbf{Value} & \textbf{Remarks} \\
\hline
Pan states & 19 & Practical deployment range of $0^{\circ}$--$180^{\circ}$ at $10^{\circ}$ resolution \\
\hline
Tilt states & 11 & Practical deployment range of $25^{\circ}$--$75^{\circ}$ at $5^{\circ}$ resolution \\
\hline
RSSI trend states & 3 & Increase, stable, and decrease \\
\hline
Actions per state & 5 & Left, right, up, down, hold \\
\hline
Total Q-values & 3135 & Compact enough for direct static allocation \\
\hline
Stored numeric type & 32-bit float & Preserves training precision with low memory overhead \\
\hline
Estimated memory footprint & 12.3~kB & Well within ESP32-S3 memory constraints \\
\hline
Inference policy & $\arg\max Q(s,a)$ & No online learning during hardware testing \\
\hline
\end{tabular}
\end{table}

The practical firmware used inference only. Online updating of Q-values was disabled to
avoid instability caused by noisy packet-level measurements, to keep computational cost
low, and to guarantee repeatable behavior between trials.

\subsubsection{Operational Sequence During Practical Trials}

During field testing, the RL-based controller was executed as a two-stage procedure.
First, the antenna was moved to a valid starting orientation and a reference RSSI batch
was acquired. Second, the embedded controller repeatedly encoded the current state,
selected the highest-valued action from the Q-table, and applied the corresponding motor
command until convergence.

The trial was considered converged when either of the following conditions was met:
\begin{itemize}
\item The absolute RSSI improvement remained below 0.8~dB for four consecutive control cycles, or
\item The algorithm reached the maximum allowed number of decision steps
\end{itemize}

This stopping rule reduced unnecessary movement near the optimum and produced consistent
trial durations for the comparison presented in the results chapter.
