\subsection{Software and Firmware Design}

\subsubsection{Firmware Structure}

The embedded software was implemented as a deterministic control loop running on the
ESP32-S3. A single-loop architecture was selected to keep the firmware predictable,
portable, and easy to validate during repeated field trials. Instead of using an
RTOS-based multi-task design, the implementation separates sensing, decision making,
and actuation as sequential software blocks executed in a fixed order.

Each control cycle performs the following operations:
\begin{itemize}
\item Acquisition of a batch of RSSI samples from the received packets
\item Averaging and filtering of the sampled RSSI values
\item Encoding of the current pan angle, tilt angle, and RSSI trend
\item Selection of the next action from the active alignment algorithm
\item Execution of stepper and servo commands
\item Logging of the experiment state for post-processing
\end{itemize}

This execution order ensured that every alignment method operated on the same measured
feedback and experienced the same settling delay between consecutive movements.


\subsubsection{RSSI Processing and State Encoding}

RSSI measurements are inherently noisy due to multipath propagation, short-term fading,
and packet-to-packet fluctuations. To reduce this variability, a small batch of samples
is collected at each orientation and averaged before the decision block is executed.
The averaged RSSI at control step $k$ is given by:
\begin{equation}
\overline{\text{RSSI}}_k = \frac{1}{N} \sum_{i=1}^{N} \text{RSSI}_{k,i}
\end{equation}
where $N$ denotes the number of samples collected at the current orientation.
The short-term RSSI trend is then estimated as:
\begin{equation}
\Delta \text{RSSI}_k = \overline{\text{RSSI}}_k - \overline{\text{RSSI}}_{k-1}
\end{equation}
The trend variable is quantized into increasing, stable, and decreasing classes using
a small dead-band threshold $\epsilon$ to suppress noise-driven state changes.

The embedded state is encoded as:
\begin{equation}
s_k = \left( i_{\text{pan}},\; i_{\text{tilt}},\; \text{trend}_k \right)
\end{equation}
where $i_{\text{pan}}$ and $i_{\text{tilt}}$ are discrete orientation indices derived from
fixed angular step sizes.

\subsubsection{Firmware Modules}

The final firmware was organized into modular functions so that the exhaustive scan,
hill-climbing method, and RL-based policy could share common sensing and actuation
primitives. This reduced implementation bias during algorithm comparison.

\vspace{0.5cm}
\begin{table}[H]
\centering
\caption{Embedded Firmware Modules}
\label{tab:firmware_modules}
\renewcommand{\arraystretch}{1.4}
\begin{tabular}{|
>{\raggedright\arraybackslash}p{3.2cm}|
>{\raggedright\arraybackslash}p{4.2cm}|
>{\raggedright\arraybackslash}p{6.2cm}|
}
\hline
\textbf{Module} & \textbf{Implemented On} & \textbf{Function} \\
\hline
RSSI Acquisition & ESP32 Wi-Fi interface & Captures packet RSSI values and stores a short measurement window for averaging \\
\hline
Motion Control & GPIO and MCPWM peripherals & Generates STEP/DIR pulses for pan motion and PWM pulses for tilt adjustment \\
\hline
State Encoder & Application layer firmware & Maps continuous angles and RSSI trend to discrete state indices \\
\hline
Policy Executor & Application layer firmware & Runs exhaustive scan, hill-climbing, or greedy Q-table inference \\
\hline
Data Logger & UART serial output & Records trial identifier, angles, RSSI, elapsed time, and convergence status \\
\hline
\end{tabular}
\end{table}

\subsubsection{Control Timing and Runtime Parameters}

A consistent timing configuration was maintained across all algorithms to ensure fair
comparison during practical evaluation. After every commanded movement, the firmware
waited for mechanical settling before the next RSSI batch was sampled. This prevented
measurement corruption caused by vibration and transient misalignment.

\vspace{0.5cm}
\begin{table}[H]
\centering
\caption{Key Runtime Parameters Used in Practical Experiments}
\label{tab:runtime_parameters}
\renewcommand{\arraystretch}{1.4}
\begin{tabular}{|
>{\raggedright\arraybackslash}p{3.2cm}|
>{\raggedright\arraybackslash}p{3.2cm}|
>{\raggedright\arraybackslash}p{7.2cm}|
}
\hline
\textbf{Parameter} & \textbf{Value} & \textbf{Remarks} \\
\hline
Pan sweep range & $0^{\circ}$ to $180^{\circ}$ & Limited to the mechanically reachable front hemisphere during trials \\
\hline
Tilt sweep range & $25^{\circ}$ to $75^{\circ}$ & Selected to cover the practical mounting geometry \\
\hline
Coarse pan step & $10^{\circ}$ & Used for exhaustive scan and initial search stages \\
\hline
Coarse tilt step & $5^{\circ}$ & Matches the servo control resolution used in testing \\
\hline
Fine step size & $2^{\circ}$ & Used for local refinement around promising states \\
\hline
Samples per point & 20 & Average computed from 20 consecutive RSSI readings \\
\hline
Sample delay & 10~ms & Delay between individual RSSI samples within a batch \\
\hline
Settling delay after movement & 180~ms & Allows mechanical oscillation to decay before measurement \\
\hline
\end{tabular}
\end{table}

\subsubsection{Experiment Logging Format}

To simplify post-processing, each practical run was stored using a structured experiment
header describing the environment and scan parameters. The recorded format followed the
same field layout throughout the project, which made the generated data compatible with
spreadsheet-based analysis and script-based aggregation.

\vspace{0.5cm}
\begin{table}[H]
\centering
\caption{Practical Experiment Log Header}
\label{tab:experiment_log_header}
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{|
>{\raggedright\arraybackslash}p{3.8cm}|
>{\raggedright\arraybackslash}p{10.2cm}|
}
\hline
\textbf{Field} & \textbf{Description} \\
\hline
Experiment\_ID & Unique identifier of the trial set \\
\hline
Environment & Indoor or outdoor deployment condition \\
\hline
Distance\_TX\_RX\_m & Separation between transmitter and receiver antennas \\
\hline
Obstacle\_Description & LOS, partial blockage, corridor wall, vegetation, and similar descriptors \\
\hline
Pan\_Min/Max, Tilt\_Min/Max & Mechanical search limits used for the run \\
\hline
Pan\_Step, Tilt\_Step, Fine\_Step & Control resolution used by the selected algorithm \\
\hline
Steps\_Per\_Degree & Conversion factor between logical angle and motor command \\
\hline
Samples\_Per\_Point, Sample\_Delay\_ms & RSSI averaging settings for each orientation \\
\hline
Notes & Additional observations recorded during the field run \\
\hline
\end{tabular}
\end{table}
